Widespread nitrous oxide undersaturation in U.S. lakes and reservoirs

### 2023-04-21 manuscript/manuscript_file.Rmd
not sure what changes are in this commitgit add -u

```{r table1, echo = FALSE}
rbind(national_stats %>% 
        select(WSA9, estimate, LCL, UCL, post_med_sat, LCI_sat, UCI_sat, post_med_en2o, LCL_en2o, UCL_en2o),
      ecoreg_stats %>% 
        select(WSA9, estimate, LCL, UCL, post_med_sat, LCI_sat, UCI_sat, post_med_en2o, LCL_en2o, UCL_en2o)) %>%
  mutate('N2O undersaturation extent' = paste0(estimate*100, " (", LCL*100, "-", UCL*100, ")"),
         'N2O saturation ratio' = paste0(post_med_sat, " (", LCI_sat, "-", UCI_sat, ")"),
         'N2O emission rate' = paste0(post_med_en2o, " (", LCL_en2o, "-", UCL_en2o, ")")) %>% 
  select(WSA9, 'N2O undersaturation extent', 'N2O saturation ratio', 'N2O emission rate') %>%
  rename('spatial domain' = WSA9) %>%
knitr::kable(caption = "Table 1. Mean and 95% confidence interval of N2O undersaturation extent, saturation ratio, and emission rate at the national and ecoregional scales")

@@

#### Controls on N~2~O
Dissolved N~2~O concentration increased with NO~3~ concentration **[need to pick a figure]**, likely because NO~3~ concentration partially controls denitrification, a type of anaerobic metabolism that couples the oxidation of organic matter with the  sequential reduction of NO~3~ to nitrite (NO~2~), nitric oxide (NO), N~2~O, and N~2~ (Knowles 1982; Richardson et al. 2009).  The reduction of NO~3~ to N~2~ can be incomplete, however, resulting in the release of N~2~O to the water column.  Incomplete denitrification often occurs when NO~3~ is relatively abundant, likely because the final reduction of N~2~O to N~2~ is the least energetic reaction in the denitrification pathway, yielding less than 20% of the total energy yield of complete denitrification (Koike and Hattori, 1975;Richardson et al., 2009). When NO~3~ is limiting, however, denitrification can resort to scavenging N~2~O from the water column and reducing it to N~2~ thereby functioning as an N~2~O sink.

@@

The sensitivity of N2O to NO3 increased with water temperature **[need to pick a figure]**.  This is consistent with N2O production and consumption occurring through a temperature sensitive biological process.  This relationship is apparent at high NO3 concentrations, where nitrogen biogeochemistry is a net source of N2O, and at low NO3 concentrations where nitrogen biogeochemistry is a net sink for N2O.  Thus NO3 concentration dictates whether a water body is a source or sink of N2O and elevated water temperature magnifies the effect.

**[N2O vs NO3 with different lines representing quantiles of the temperature distribution]**

#### Controls on N~2~O saturation **[this heading is to organize the .Rmd, not intended to be retained in final MS]**

While NO3 availability determined whether N biogeochemistry was a source or sink of N2O and water temperature affected the rate of N2O production or consumption, waterbody morphology determined the extent to which N2O production or consumption affected dissolved N~2~O concentration at the air-water interface.  When pooled across all ecoregions, the absolute magnitude of the difference between the measured and equilibrium N~2~O concentration, known as $\Delta$ N~2~O, decreases linearly with increasing waterbody size class (Figure 3A). Waterbody size can affect dissolved gas concentration through several mechanisms. In lakes with low volume to benthic surface area ratios, the mass of N~2~O consumed or produced in the sediments can be a substantial fraction of the mass of N~2~O dissolved in the water column.  Since water volume to benthic surface area ratios tend to scale positively with lake surface area, benthic N~2~O biogeochemistry will have a greater affect on dissolved N~2~O concentration in small lakes, regardless of whether the sediments function as a net sink or source of N~2~O **[see Wetzel for references]**.  Furthermore, sediment may warm faster in small and shallow lakes, whereas thermal stratification may substantially delay sediment warming in large and deep lakes, resulting in greater rates of N~2~O production or consumption in small lakes **[need a citation for N2O ~ temperature relationship]**.  Thermal stratification may also prevent the exchange of water between the lake surface and the hypolimnion, thereby preventing N~2~O depleted or enriched water from advecting to the lake surface (Beaulieu et al N2O Harsha Lake and USACE N2O papers).

@@

The relationship between N~2~O and waterbody size is also evident in a few ecoregions (Figure 3B).  For example, mean N2O saturation ratio is greater than 1 for all size classed in the Coastal Plains, consistent with the high NO3 concentrations in this ecoregion **[is this true?]**, but decreases with increasing lake size, consistent the effect of waterbody morphology on N2O production and gas exchange.  By contrast, mean N2O saturation is less than one across all size classes in the low NO3 northern plains, but again, N2O saturation moves toward equilibrium with increasing lake size.  These relationships are evident to lesser degrees in other ecoregions, likely dependent on the distribution on NO3 and water temperature across lake classes.  **[I'm beginning to think that Fig. 3B unnecessarily complicates an otherwise pretty clean story.  Maybe we just show Fig. 3B?]**

@@

# EMISSION RATES

>>>>

```{r table1, echo = FALSE}
rbind(national_stats %>% 
        select(WSA9, estimate, LCL, UCL, post_med_sat, LCI_sat, UCI_sat, post_med_en2o, LCL_en2o, UCL_en2o),
      ecoreg_stats %>% 
        select(WSA9, estimate, LCL, UCL, post_med_sat, LCI_sat, UCI_sat, post_med_en2o, LCL_en2o, UCL_en2o)) %>%
  mutate('N2O undersaturation extent' = paste0(estimate*100, " (", LCL*100, "-", UCL*100, ")"),
         'N2O saturation ratio' = paste0(post_med_sat, " (", LCI_sat, "-", UCI_sat, ")"),
         'N2O emission rate' = paste0(round(post_med_en2o, digits = 1), " (", round(LCL_en2o, digits = 1), "-", round(UCL_en2o, digits = 1), ")")) %>% 
  select(WSA9, 'N2O undersaturation extent', 'N2O saturation ratio', 'N2O emission rate') %>%
  rename('spatial domain' = WSA9) %>%
knitr::kable(caption = "Table 1. Mean and 95% confidence interval of N2O undersaturation extent, saturation ratio, and emission rate at the national and ecoregional scales")

@@

#### Controls on N~2~O
##### NO3
Dissolved N~2~O concentration increased with NO~3~ concentration **[need to pick a figure]**, likely because NO~3~ concentration partially controls denitrification, a type of anaerobic metabolism that couples the oxidation of organic matter with the  sequential reduction of NO~3~ to nitrite (NO~2~), nitric oxide (NO), N~2~O, and N~2~ (Knowles 1982; Richardson et al. 2009).  The reduction of NO~3~ to N~2~ can be incomplete, however, resulting in the release of N~2~O to the water column.  Incomplete denitrification often occurs when NO~3~ is relatively abundant (Beaulieu et al Harsha Lake, Beauliue et al USACE), likely because the final reduction of N~2~O to N~2~ is the least energetic reaction in the denitrification pathway, yielding less than 20% of the total energy yield of complete denitrification (Koike and Hattori, 1975;Richardson et al., 2009). When NO~3~ is limiting, however, denitrification can resort to scavenging N~2~O from the water column and reducing it to N~2~, thereby functioning as an N~2~O sink.

@@

##### Temperature
The sensitivity of N2O to NO3 increased with water temperature **[need to pick a figure]**, which is consistent with N2O production and consumption occurring through temperature sensitive biological processes.  At low NO3 concentrations, greater water temperature exacerbated N2O undersaturation, presumably by stimulating N2O consumption rates.  At high NO3 concentration, greater water temperature increased N2O supersaturation, likely by stimulating N2O production rates.  The relative effect of NO3 was greater than that of temperature, thus elevated water temperature serves to magnify the rates of N2O production or consumption dictated by NO3 availability.

##### Waterbody size

While NO3 and water temperature were the primary drivers of N2O production or consumption rates, waterbody morphology determined the extent to which N2O processing affected dissolved N~2~O in surface waters.  When pooled across all ecoregions, the absolute magnitude of the difference between the measured and equilibrium N~2~O concentration, known as $\Delta$ N~2~O, decreases with increasing waterbody size class (Figure 3A). Waterbody size can affect dissolved gas concentration through several mechanisms. In lakes with low volume to benthic surface area ratios, the mass of N~2~O consumed or produced in the sediments can be a substantial fraction of the mass of N~2~O dissolved in the water column.  Since water volume to benthic surface area ratios tend to scale positively with lake surface area, benthic N~2~O biogeochemistry will have a greater affect on dissolved N~2~O concentration in small lakes, regardless of whether the sediments function as a net sink or source of N~2~O **[see Wetzel for references]**.  Furthermore, sediment may warm faster in small and shallow lakes, whereas thermal stratification may substantially delay sediment warming in large and deep lakes, resulting in greater rates of N~2~O production or consumption in small lakes **[need a citation for N2O ~ temperature relationship]**.  Thermal stratification may also prevent the exchange of water between the lake surface and the hypolimnion, thereby preventing N~2~O depleted or enriched water from advecting to the lake surface (Beaulieu et al N2O Harsha Lake and USACE N2O papers).

@@

The influence of waterbody size is also evident in the statistical interaction between NO3 and surface area (Fig XX).  At high NO3 concentrations, N2O is highly supersatured in small waterbodies, but moves toward equilibrium in large waterbodies, as seen in the Coastal Plains ecoregion**[is CPL high NO3?]**.  At low NO3 concentrations, N2O is undersatured in small waterbodies, but increases toward equilibrium in larger waterbodies, as seen in the Northern Plains ecoregion **[is NPL low nitrate?]**(Fig).

The model indicates that NO3, water temperature, and surface area interact to determine the N2O saturation ratio.  These interactions yield a broad range of N2O saturation values across the population of CONUS waterbodies, but also enable the identifcation of broad scale patterns.  First, small waterbodies with low NO3 concentration tend to be undersaturated, whereas small waterbodies with high NO3 tend to be highly supersaturated.  Large waterbodies, regardless of NO3 status, tend to have N2O saturation ratios close to equilibrium.  Therefore intra and inter regional patterns in N2O saturation are determined by the joint distributions of waterbody size, NO3 availability, and water temperature within an ecoregion.

@@

# EMISSION RATES
Despite widespread N2O undersaturation in CONUS waterbodies, the national average N2O emission rate was `r national_stats$post_med_en2o` (95% CI: `r national_stats$LCL_en2o` - `r national_stats$UCL_en2o` mg N2O m^-2^ d^-1^ and only the Northern Appalachians and Northern Plains ecoregions had mean emission rates less than 0 (Table 1).  This is accordance with the distribution on N2O saturation values where relatively rare, but highly supersatured, waterbodies pull the mean to a value that far exceeds the median.    ecoregional means range from `r ecoreg_stats %>% slice_min(n = 1, post_med_en2o) %>% select(post_med_en2o) %>% pull()` to `r ecoreg_stats %>% slice_max(n = 1, post_med_en2o) %>% select(post_med_en2o) %>% pull()` mg N2O m^-2^ d^-1^ (Table 1).
